Anomalous waves triggered by abrupt depth changes: laboratory experiments and truncated KdV statistical mechanics

### 2019-11-16 paper2_JNLS/paper.tex
NM

Our particular focus is the laboratory experiments of \boetal~\cite{bolles2019}, who demonstrated the emergence of anomalous statistics from a randomized wave-field encountering an abrupt depth change (ADC). In these experiments, the incoming wave-field is generated with nearly Gaussian statistics, and a plexiglass step placed near the middle of the tank creates the depth change. Upon passing into the shallower region, the wave distribution skews strongly towards positive displacement, with deviation from Gaussian being most pronounced a short distance downstream of the ADC. Inspired by these experimental observations, \maetal~\cite{majda2019} developed a theoretical framework that accurately captures several key aspects of the anomalous behavior. The theory is based on deterministic and statistical analysis of the truncated Korteweg–de Vries (TKdV) equations, and uses a combination of computational, statistical, and analytical tools. In subsequent work, Majda and Qi analyzed more severely truncated systems --- as low as two interacting modes --- that exhibit a statistical phase transition to anomalous statistics and the creation of extreme events while enjoying a more tractable structure \cite{majdaqi2019}. More recently, Qi and Majda demonstrated the capability of machine learning strategies to accurately predict these extreme events \cite{MachineLearning2019}. That work uses a deep neural network coupled with a judicious choice of the entropy loss function that emphasizes the dominant structures of the turbulent field \cite{MachineLearning2019}. It should be noted that, in these studies, extreme events are represented by strong skewness of the wave-field. They occur intermittently and on a relatively frequent basis, which is in contrast to the simpler situation of isolated, rare events \cite{guth2019machine}.

The purpose of the present manuscript is to provide a more comprehensive treatment of the {\em combined} experimental and theoretical efforts of \boetal~\cite{bolles2019} and \maetal~\cite{majda2019, majdaqi2019}. Special emphasis is placed on the synergy between experiment and theory -- a cooperative strategy with a proven record of success \cite{camassa2012stratified, ristroph2012, ganedi2018equilibrium}. 
In Section \ref{experiments} we detail the laboratory experiments of \boetal~and summarize previously reported findings on anomalous statistics of the surface displacement. In this section, we also present new experimental data on surface {\em slope} statistics --- a line of inquiry motivated by the appearance of the slope in the Hamiltonian structure of TKdV theory developed by \maetal~\cite{majda2019}.
In Section \ref{theory}, we detail the TKdV theoretical framework, with a more thorough treatment of the link between model and experimental parameters than was reported in previous work. This link facilitates comparison between the theory and experiments. Within the same TKdV framework, we discuss both a deterministic dynamical systems approach and a statistical mechanics approach. The latter is based on the Hamiltonian structure of TKdV and novel ensemble distributions that incorporate canonical and micro-canonical aspects.
Section \ref{results} presents a systematic comparison between experimental measurements and model predictions, with both surface displacement and surface slope statistics compared closely. This section, also describes deterministic dynamics of TKdV, which offer insight into the experimental length and timescales required for wave statistics to transition. Importantly, this section shows newly collected experimental data related to a a peculiar relationship between surface-displacement skewness and surface-slope variance that is predicted by the theory. The experimental data provides conclusive confirmation of the model prediction. We close with some final remarks in Section \ref{conclusion}.

%confirms the predictive power of the framework developed by \maetal~\cite{majda2019}. 
% spectacular, remarkable, marvelous, conclusively, convincingly

%We first provide a more thorough treatment of the link between experimental and model parameters in order to facilitate a detailed comparison between the two. Next, we present new experimental data, most notably on {\em surface slope} statistics, which motivates parallel examination of the theory.  Likewise, novel theoretical predictions have inspired us to perform additional experimental measurements, especially as it pertains to a peculiar relationship between surface-displacement skewness and surface-slope variance that is predicted by the theory. Comparison with the new measurements further confirms the predictive power of the framework developed by \maetal~\cite{majda2019}.

% NOTES
% Words: conjunction, combination, close cooperation; elucidate.
% New data: slopes, autocorrelation, depth ratio varied.

>>>>

Our particular focus is the laboratory experiments of \boetal~\cite{bolles2019}, who demonstrated the emergence of anomalous statistics from a randomized wave-field encountering an abrupt depth change (ADC). In these experiments, the incoming wave-field is generated with nearly Gaussian statistics, and a plexiglass step placed near the middle of the tank creates the depth change. Upon passing over the step, the wave distribution skews strongly towards positive displacement, with deviation from Gaussian being most pronounced a short distance downstream of the ADC. Inspired by these experimental observations, \maetal \cite{majda2019} developed a theoretical framework that accurately captures several key aspects of the anomalous behavior. The theory is based on deterministic and statistical analysis of the truncated Korteweg–de Vries (TKdV) equations, and uses a combination of computational, statistical, and analytical tools. In subsequent work, Majda \& Qi (2019) \cite{majdaqi2019} analyzed more severely truncated systems --- as low as two interacting modes --- that exhibit a statistical phase transition to anomalous statistics and the creation of extreme events while enjoying a more tractable structure. More recently, Qi \& Majda (2019) \cite{MachineLearning2019} demonstrated the capability of machine learning strategies to accurately predict these extreme events. That work uses a deep neural network coupled with a judicious choice of the entropy loss function that emphasizes the dominant structures of the turbulent field \cite{MachineLearning2019}. It should be noted that, in these studies, extreme events are represented by strong skewness of the wave-field. They occur intermittently and on a relatively frequent basis, which is in contrast to the simpler situation of isolated, rare events \cite{guth2019machine}.

The purpose of the present manuscript is to provide a more comprehensive treatment of the {\em combined} experimental and theoretical efforts of \boetal~\cite{bolles2019} and \maetal~\cite{majda2019, majdaqi2019}. Special emphasis is placed on the synergy between experiment and theory -- a cooperative strategy with a proven record of success \cite{camassa2012stratified, ristroph2012, ganedi2018equilibrium}. In Section \ref{experiments} we detail the laboratory experiments of \boetal \cite{bolles2019} and summarize previously reported findings on anomalous statistics of the surface displacement. We also present new experimental data on surface {\em slope} statistics --- a line of inquiry motivated by the theoretical advancements of \maetal \cite{majda2019}, specifically the central role played by the slope in the Hamiltonian structure of TKdV.  Section \ref{theory} discusses the TKdV theoretical framework, including both the deterministic and the statistical mechanics perspectives. In this section, we flesh out details of the non-dimensionalization, which, for brevity, were only briefly discussed in previous work. This exercise provides a more thorough link between model and experimental parameters, ultimately facilitating a closer comparison between the two. We then discuss TKdV as a deterministic dynamical system, approachable from the viewpoint of statistical mechanics. The later perspective is based on the Hamiltonian structure and novel ensemble distributions that incorporate canonical and micro-canonical aspects \cite{abramov2003}. Section \ref{results} presents a systematic comparison between experiments and theory, focusing on both surface displacement and surface slope. We find the statistical distributions of these two quantities to agree well across experiments and theory. In particular, we find the agreement in displacement statistics to be remarkably quantitative. We then examine numerical simulations of the TKdV deterministic dynamics, which elucidates the length and timescales required for statistical transitions in the experiments. Finally, this section discusses a peculiar power-law relationship between surface-displacement skewness and surface-slope variance predicted by the statistical mechanics framework \cite{majda2019}. New experimental measurements conclusively confirm this prediction, further demonstrating the predictive power of the TKdV framework developed by \maetal \cite{majda2019}. We close with some final remarks in Section \ref{conclusion}.
